%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}


% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{proof}
\usepackage{float} 
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{todonotes}
\usepackage{comment}
\usepackage{tocbibind}
\usepackage[toc,page]{appendix}
\usepackage{lipsum}   




\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subfigure}
\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1.0,1.0,1.0}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
 commentstyle=\color{mygreen},    % comment style
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  keywordstyle=\color{blue},       % keyword style
  stringstyle=\color{mymauve}    % string literal style
}

\lstset{style=mystyle}

\usepackage{setspace}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}


\newcommand{\li}[1]{\lstinline{#1}}
\newcommand{\keywadj}[1]{\mathtt{#1}}
\newcommand{\keyw}[1]{\keywadj{#1}~}
\newcommand{\m}[1]{\ \texttt{#1}\ }
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}{Definition}[section]



% Provides a draft mark at the top of the document. 
\draftstamp{\today}{DRAFT}

\begin {document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Extending Abstract Effects with Bounds and Algebraic Handlers}}
\author{Anlun Xu}
\date{November 2020}
\Year{2020}
\trnumber{}

\committee{
\ 
}

\support{}
\disclaimer{}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{Effect Systems, Algebraic Effects and Handlers}

\maketitle

\begin{dedication}
Dedication
\end{dedication}

\pagestyle{plain} % for toc, was empty


\begin{abstract}
Effect systems have been a subject of active research for nearly four decades, with the most notable practical example being checked exceptions in programming languages such as Java.  The work on effects can be divided into two strands: The \textit{restrictive} approach (e.g., Java's checked exceptions), which tracks effects that are already built into the language--such as reading and writing state or exceptions--and provides a way to restrict them. And the \textit{denotational} approach, which includes algebraic effects, defines the semantics of computational effects based on primitives. While there are many existing restrictive or denotational effect systems, they are rarely designed with scalability in mind. In this thesis, we design multiple effect systems around the idea of making effect system scalable when developing large and complex software. The first part of our work is a restrictive path-dependent effect system that provides a granular effect hierarchy by allowing abstract effect members to be bounded. This thesis presents a full formalization of the effect-system, and provides an implementation as a part of the Wyvern programming language. The second part of our work presents a denotational effect-system that supports abstract algebraic effects. This thesis gives a formalization of the system and provides proofs for type soundness and correctness of dynamic semantics. 
\end{abstract}
\begin{acknowledgments}
\end{acknowledgments}



\tableofcontents
\listoffigures

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...

\input{Introduction}
\input{background}
\input{bounded}
\input{algebraic}
\input{module}
\input{conclusion}


%\appendix
%\include{appendix}

\backmatter

%\renewcommand{\baselinestretch}{1.0}\normalsize

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat}
\bibliography{related} %your bib file

\appendix
\input{appendix}


\end{document}
